Computational methods in landscape ecology

### 2024-10-17 Sections/04-connectivity.qmd
further minor changes

<!-- R1: Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g. animals, gravity, wind, or water -->
<!-- R1: facilitate or impeded movement and flow -->
Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms, e.g., active movement of animals, or dispersal of sessile plants by wind or water, and can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Landscape metrics are commonly used to describe structural connectivity in terms of e.g., total habitat amount, patch size and isolation, or possible connections between habitat patches.
Recently, there has been an increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity [@Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

@@

Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
<!-- R1: to traverse or disperse through that cell -->
A resistance surface is a raster-based representation of a landscape where each cell is assigned a value representing the cost for an individual of a particular species to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
<!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
Creating resistance surfaces involves obtaining landscape data for the area of interest, quantifying cost values for each cell using movement and dispersal data, and finally analyzing the surfaces [@Dutta2022OverviewComputational].
This can include expert opinion (widely used due to low effort, however, difficult to measure accuracy) [@Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).

<!-- R!: Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything. Assumptions are rather in how such results are used in an ecological context -->
Least-cost modelling identifies potential paths between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and related connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, graph theory based metrics can assist wildlife movement corridor planning [@Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].

Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement and dispersal between patches is modeled as edges.
<!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
The edges can be binary representing whether a connection between two nodes either exists or not, weighted by cost values representing movement efforts from one patch to another, or undirected or directed allowing movement and dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework was initially only applied to single species but is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches enable the modelling of not only spatial, but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or from different data sources.
<!-- R1: I don't think that it is specific to graphs. Any data-dependent model will be sensitive to input data. This highlights the importance of data -->
In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [@Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

>>>>

<!-- R1: Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g. animals, gravity, wind, or water -->
<!-- R1: facilitate or impeded movement and flow -->
Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms (e.g., active movement of animals, or dispersal of sessile plants by wind or water). 
It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
Connectivity can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Landscape metrics are commonly used to describe structural connectivity in terms of, e.g., total habitat amount, patch size, or patch isolation.
Recently, there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity [@Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

@@

Many modern connectivity approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
<!-- R1: to traverse or disperse through that cell -->
A resistance surface is a raster-based representation of a landscape where each cell is assigned a value reflecting the species-specific cost for an individual to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
<!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
Creating resistance surfaces involves obtaining landscape data for the area of interest, quantifying cost values for each cell using movement and dispersal data, and finally analyzing the surfaces [@Dutta2022OverviewComputational].
This can include expert opinion (widely used due to low effort, however, difficult to measure accuracy) [@Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).

<!-- R!: Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything. Assumptions are rather in how such results are used in an ecological context -->
Least-cost modelling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
By calculating pathways between two points it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement and dispersal in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and related connectivity metrics can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, graph theory based metrics can assist wildlife movement corridor planning [@Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].

Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement and dispersal between patches is modeled as edges.
<!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
The edges can be binary representing whether a connection between two nodes exists or not, weighted by cost values representing movement or dispersal efforts from one patch to another, or undirected or directed allowing movement or dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity which is based on a binary connection model [@Pascual-Hortal2006ComparisonDevelopment; @Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and species-specific movement or dispersal [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches enable the modelling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or various data sources.
<!-- R1: I don't think that it is specific to graphs. Any data-dependent model will be sensitive to input data. This highlights the importance of data -->
In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [@Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

### 2024-10-17 Sections/05-sampling.qmd
further minor changes

Establishing a pattern-process link usually requires fieldwork and data collection.
Given the cost of collecting field observations across landscapes computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
To improve the statistical significance of the relationships between landscape features and field observations several aspects need to be considered when selecting sampling sites for data collection.
First, selected sites should cover the maximum possible range of landscape heterogeneity to maximize the variance of response variables.
Second, statistical power can be improved by ensuring independence between sites and consequently response variables.
Finally, since many landscape features that influence response variables are spatially dependent it is essential to evaluate spatial autocorrelation.

@@

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures (e.g., Moran's I) between sampling approaches or spatial scales [@Pasher2013OptimizingLandscape].
Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

>>>>

Establishing a pattern-process link usually requires fieldwork and data collection.
Given the cost of collecting field observations across landscapes, computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
To improve the statistical significance of the relationships between landscape features and field observations several aspects need to be considered when selecting sampling sites for data collection.
First, selected sites should cover the maximum possible range of landscape heterogeneity to maximize the variance of response variables.
Second, statistical power can be improved by ensuring independence between sites and consequently response variables.
Finally, since many landscape features that influence response variables are spatially dependent, it is essential to evaluate spatial autocorrelation.

@@

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures between sampling approaches or spatial scales (e.g., Moran's I) [@Pasher2013OptimizingLandscape].
Furthermore, spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

### 2024-10-17 Sections/06-simulations.qmd
further minor changes

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses which would be impossible in natural systems [@Peck2004SimulationExperiment].
Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we focus mainly on exploratory models but include both pattern- and process-based models.

@@

Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral] or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware]. -->

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes (i.e., landscape generators) [@Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators have allowed to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrate different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
Thus, digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].

>>>>

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, simulations can be seen as experimental systems that allow all imaginable manipulations which would be impossible in natural systems in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow to control certain aspects of the landscape using the raster data model [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware]. -->

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes (i.e., landscape generators) [@Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent, landscape generators have allowed to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands (combining pattern- and process-bases approaches) [@Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins]. 
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
